---
title: "Example workflow"
author: "James Campbell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The following example will show you the typical workflow for downloading data from the North American Breeding Bird Survey ftp server.  First, we'll need to find the relevant codes for our species and regions of interest.  In the following examples, we'll download all data from multiple species from the United States and Canada in the years 2011-2013.

### Pull North American Breeding Bird Survey data

Lets start by pulling the meta data to get the codes.  The `pullBBSmeta` function will create a folder in the current directory in which it'll save all the meta files.

```{r,message=FALSE,warning=F}
require(pullBBS)
pullBBSmeta()
```

Check the file `BBS_Meta\RegionCodes.txt` and `BBS_Meta\SpeciesList.txt` to find the codes you need.  From browsing the files, we found these codes for our regions and species:

- Ovenbird: **06740**
- Tennasee Warbler: **06470**
- Red-breasted Nuthatch: **07280**
- Chipping Sparrow: **05600**
- Canada, United States: **124**, **840**
- BC, Alberta, Saskatchewan, Manitoba, Ontario: **11**, **04**, **79**, **45**, **68**

Next, we'll use this to pull the route info from the ftp server.

```{r,message=FALSE,warning=FALSE}
routes <- pullBBS(year = c(2011,2012,2013),country = c(124),region = c(11,04,79,45,68),
                  AOU = c(06740,06470,07280),useCache = T)
head(routes)
```

As seen above, the BBS data is in a wide format, with each of the 50 stops holding a column in the table reporting the number of observations for that stop on a particular route.  To make the data easier to handle in R, we can condense the 50 stop columns into 2 columns: `Stop` and `Observations`.  The function `WideToLongBBS` will take care of this for us.

```{r,warning=F,message=F}
routes.long <- WideToLongBBS(routes = routes)
head(routes.long)
```

### Plot the Summed Stop Observations per Route

Now we can plot our BBS route locations in `leaflet`.  Here we'll plot the summed sightings per route per species for a specified year.  Each route in the BBS has a unique lat/long coordinate and within each route there are 50 stops where the observer counts all audible species.  Ideally, each route is sampled once per year but many sites cannot be reached on a yearly basis.

On the following plot, you can examine each species by selecting the layer corresponding to its AOU code.

```{r,warning=FALSE,message=F,fig.width=7,fig.height=5}
plotBBS(routes = routes.long,year = 2011,markerSize = 4)
```

### Additiontal Examples

We'll now plot all data for a given species accross the entire United States.  Regions (states or provinces) which have 0 sightings for the specified species will be automatically excluded from the data set by the `pullBBS` function.

```{r,message=FALSE,warning=F}
routes <- pullBBS(year = c(2011),country = c(840),
                  AOU = c(06740,05600),useCache = T)
routes.long <- WideToLongBBS(routes = routes)
```

```{r,warning=FALSE,message=F,fig.width=7,fig.height=5}
plotBBS(routes = routes.long,markerSize = 2)
```

We can also pull data for all observed species within a specified region and period and then plot the first 8 species.  Setting `useCache` to `F` will wipe all temporary BBS files from your hard drive, so its usefull to set this only at the end of your workflow as having the cache will greatly speed up multiple calls to `pullBBS`.

```{r,message=FALSE,warning=F}
routes <- pullBBS(year = c(2011),country = c(124),
                  region = 04, useCache = F)
routes.long <- WideToLongBBS(routes = routes)
```

```{r}
AOUs <-unique(routes.long$AOU)
message(length(AOUs),' species found within this region.')
```

```{r,warning=FALSE,message=F,fig.width=7,fig.height=5}
plotBBS(routes = routes.long,year = 2011,markerSize = 4,AOU = AOUs[1:8])
```
